Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-02-01 causal-bandit/icml-2015/setup.tex
Moved parallel to S3. Re-did Algorithm 1 using a instead of (i,j).

We now introduce a novel class of stochastic sequential decision problems which we call \emph{causal bandit problems}. 
In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}. 
Following the terminology and notation in~\cite{Koller2009}, a \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X} = \{ X_1, \ldots, X_N \}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
An edge from variable $X_i$ to $X_j$ is interpreted to mean that a change in the value of $X_i$ may directly cause a change to the value of $X_j$.
The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$ or just $\parents{i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from the parents of $X_i$ to $X_i$ for each $X_i \in \vec{X}$. 
The resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$. 
Details can be found in Chapter 21 of~\cite{Koller2009}.

A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$, a set of \emph{observable} variables $\mathcal{O} \subseteq \mathcal{X}$, a set of \emph{allowed interventions} $\mathcal{A}$, a \emph{reward variable} $Y$. 
The following game is then played over $T$ rounds.
In round $t$, the learner selects an allowed intervention $do(\vec{X}_t = \vec{x}_t)$ from $\mathcal{A}$.
Values for all non-intervened variables $\vec{X}^c_t$ are sampled from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$. Finally, the learner is shown the sampled values of all the observable variables in $\vec{X}^c_t \cap \mathcal{O}$ and receives the reward $Y_t \in \set{0,1}$.

Although we will focus on the intervene-observe-reward ordering of events within each round, others are also possible. 
If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already well understood~\citep{Agarwal2014}. 
Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of 
allowable interventions thereby reducing the complexity.

@@

\subsection{The Parallel Bandit Problem}

We now consider a class of a causal bandit problems which we call \emph{parallel bandits}.
It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro}, including the farming example. It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.

The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \R$, as shown in Figure~\ref{fig:causalStructure}.
All variables are observable (\ie, $\mathcal{O} = \{Y, X_1, \ldots, X_N\}$) and the set of allowable actions are all size 0 and size 1 interventions: 
\[
	\mathcal{A} = \{do()\} \cup \{ do(X_i = j) \colon i \in \{1, \ldots, N\}, j \in \{0,1\}\}.
\]
In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high). 
In this case, the interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ might represent the use of shades or heat lamps to keep the temperature low or high, respectively.

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]

 %nodes
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, below right=of 2](5){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (5)
    	(2) edge (5)
    (4) edge (5);

\end{tikzpicture}
\caption{Causal model for the parallel bandits problem.\label{fig:causalStructure}}
\end{figure}

In each round of the game, the learner either purely observes by selecting $do()$ or chooses a single variable to intervene on and the value to set it to. 
The remainder of the variables are simultaneously set by independently biased coin flips. 
The value of all variables are then used to determine the distribution of rewards for that round.
Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
The value of the reward variable is distributed as $Y = r(\vec{X}) + \eta$ where $r : \{0,1\}^N \to \R$ is an arbitrary, fixed, and unknown function and $\eta$ is $1$-subgaussian noise with a distribution that may depend on $\vec{X}$.
In the farming example, this choice of $Y$ models the fixed, noisy but unknown dependence of crop yield on the temperature, soil, and moisture.

In our analysis below, we will often use the pair $(i,j)$ for $i \in \{1, \ldots, N\}$ and $j \in \{0,1\}$ as a shorthand for the intervention $do(X_i = j)$.
We will also make use of the expansion of the expected reward for $do(X_i = j)$:
\eq{
\mu_{i,j} 
&= \E{r(X)|do(X_i = j)} \\
&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k}\,.  
}
The sum and product use the restrictions $x_i = j$ and $k \ne i$ because the intervention forces $X_i = j$ with probability 1.
The optimal intervention is denoted $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$. 
Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.

@@

This enables us to estimate the the rewards for multiple interventions simultaneously by observing the values taken by the variables when we select the $do()$ action. Unfortunately interventions that occur with low probability ($\P{X_{t,i} = j}$ is small) suffer from a high approximation error and need to be explored separately, by explicitly making the intervention.

>>>>

We now introduce a novel class of stochastic sequential decision problems which we call \emph{causal bandit problems}. 
In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}. 
Following the terminology and notation in~\cite{Koller2009}, a \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X} = \{ X_1, \ldots, X_N \}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
We will assume each variable only takes on a finite number of distinct values.
An edge from variable $X_i$ to $X_j$ is interpreted to mean that a change in the value of $X_i$ may directly cause a change to the value of $X_j$.
The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$ or just $\parents{i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from the parents of $X_i$ to $X_i$ for each $X_i \in \vec{X}$. 
The resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$. 
Details can be found in Chapter 21 of~\cite{Koller2009}.

A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$ and a set of \emph{allowed interventions} $\mathcal{A}$.
One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
The following game is then played over $T$ rounds.
In round $t$, the learner \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ then \emph{observes} sampled values for all non-intervened variables $\vec{X}^c_t$ drawn from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$, including the \emph{reward} $Y_t \in \{0,1\}$.

Although we will focus on the intervene-then-observe ordering of events within each round, other scenarios are also possible. 
If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already well understood~\citep{Agarwal2014}. 
Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of 
allowable interventions thereby reducing the complexity.

@@

@@

This enables us to estimate the rewards for multiple interventions simultaneously by observing the values taken by the variables when we select the $do()$ action. Unfortunately interventions that occur with low probability ($\P{X_{t,i} = j}$ is small) suffer from a high approximation error and need to be explored separately, by explicitly making the intervention.

### 2016-02-01 causal-bandit/icml-2015/simple-regret.tex
Removed line 8
Moved parallel to S3. Re-did Algorithm 1 using a instead of (i,j).

In this section we develop and analyse an algorithm for minimising the simple regret in the
causal bandit problem.

\begin{algorithm}[h]
\caption{Causal Best Arm Identification}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} $T, N$
\FOR{$t \in 1,\ldots,(T - 1) / 2$}
\STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
\ENDFOR
\STATE Estimate $\mu$ using observation data:
\eq{
(\forall i,j) \qquad \hat \mu_{i,j} = \frac{\sum_{t=1}^{T/2} \ind{X_{t,i} = j} r_t}{\sum_{t=1}^{T/2} \ind{X_{t,i} = j}}\,.
}
\STATE Compute $\hat V_{ij} = \frac{T}{2 \sum_{t=1}^{T/2}\ind{X_{t,i}=j}} $
\STATE Compute $\hat m = m(\hat{\boldsymbol{V}})$
\STATE $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
\STATE Compute $A$ as the subset of infrequently observed arms $\set{(i,j):V_{ij} \geq m}$ with $|A| = \hat m$

\FOR{$(i,j) \in A$}
\FOR{$t \in 1,\ldots,\frac{T - 1)}{2\hat m}$}
\STATE Choose action $do(X_{t,i} = j)$ and observe $r_t$
\ENDFOR
\STATE Recompute $\hat \mu_{i,j} = \frac{2\hat m}{T} \sum_{t=1}^{ T/2\hat m} r_t(X_{t,i}=j)$ 
\ENDFOR
\STATE Estimated optimal action is $\hat i^*, \hat j^* = \argmax_{i,j} \hat \mu_{i,j}$
\STATE Choose action $do(X_{t,\hat i^*} = \hat j^*)$
\end{algorithmic}
\end{algorithm}

>>>>

In this section we develop an algorithm for achieving the optimal simple regret in a special case of a causal bandit problem.

\subsection{The Parallel Bandit Problem}

We now consider a class of a causal bandit problems which we call \emph{parallel bandits}.
It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro}, including the farming example. It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.

The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \R$, as shown in Figure~\ref{fig:causalStructure}.
All variables are observable (\ie, $\mathcal{O} = \{Y, X_1, \ldots, X_N\}$) and the set of allowable actions are all size 0 and size 1 interventions: 
\[
	\mathcal{A} = \{do()\} \cup \{ do(X_i = j) \colon i \in \{1, \ldots, N\}, j \in \{0,1\}\}.
\]
In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high). 
In this case, the interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ might represent the use of shades or heat lamps to keep the temperature low or high, respectively.

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]

 %nodes
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, below right=of 2](5){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (5)
    	(2) edge (5)
    (4) edge (5);

\end{tikzpicture}
\caption{Causal model for the parallel bandits problem.\label{fig:causalStructure}}
\end{figure}

In each round of the game, the learner either purely observes by selecting $do()$ or chooses a single variable to intervene on and the value to set it to. 
The remainder of the variables are simultaneously set by independently biased coin flips. 
The value of all variables are then used to determine the distribution of rewards for that round.
Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
\todom{Should $Y$ be 0-1 valued or real everywhere?}
The value of the reward variable is distributed as $Y = r(\vec{X}) + \eta$ where $r : \{0,1\}^N \to \R$ is an arbitrary, fixed, and unknown function and $\eta$ is $1$-subgaussian noise with a distribution that may depend on $\vec{X}$.
In the farming example, this choice of $Y$ models the fixed, noisy but unknown dependence of crop yield on the temperature, soil, and moisture.

\todom{$i,j$ notation to $a \in \actions$}
In our analysis below, we will often use the pair $(i,j)$ for $i \in \{1, \ldots, N\}$ and $j \in \{0,1\}$ as a shorthand for the intervention $do(X_i = j)$.
We will also make use of the expansion of the expected reward for $do(X_i = j)$:
\eq{
\mu_{i,j} 
&= \E{r(X)|do(X_i = j)} \\
&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k}\,.  
}
The sum and product use the restrictions $x_i = j$ and $k \ne i$ because the intervention forces $X_i = j$ with probability 1.
The optimal intervention is denoted $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$. 
Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.

\subsection{The Parallel Bandit Algorithm}
\label{sub:par-bandit-alg}

\begin{algorithm}[h]
\caption{Causal Best Arm Identification}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} Total rounds $T$, causal graph $\mathcal{G}$, allowable interventions $\actions$. 
\FOR{$t \in 1,\ldots,(T - 1) / 2$}
\STATE Perform empty intervention $do()$; observe $\vec{x}_t$ and $r_t$
\ENDFOR
\FOR{$a = do(X_i = x) \in \actions$}
\STATE Count times $a$ was seen $T_a = \sum_{t=1}^{T/2} \ind{x_{t,i} = x}$
\STATE Estimate reward $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{x_{t,i} = x} r_t$
\STATE Estimate probabilities $\hat{p}_a = \frac{2 T_a}{T-1}$ and $\hat{V}_a = \frac{1}{\hat{p}_a}$
\ENDFOR
\STATE Compute $\hat{m} = m(\hat{\vec{V}})$ and $A = \set{a \in \actions \colon \hat{V}_a \geq \hat{m}}$.
\STATE Let $T_A := \frac{T-1}{2 |A|}$ be times to sample each $a\in\actions$.
\FOR{$a = do(X_i = x) \in A$}
\FOR{$t \in 1,\ldots,T_A$}
\STATE Intervene with $a$ and observe $r_t$
\ENDFOR
\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} r_t$
\ENDFOR
\RETURN estimated optimal $\hat{a}^* \in \argmax_{a\in\actions} \hat{\mu}_a$
\end{algorithmic}
\end{algorithm}
